---
title: "Atlantic Hurricane Tracks: an exploration of HURDAT2"
author: "leon5354"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
    toc: yes
    toc_depth: 2
urlcolor: blue
---

```{r setup, include=FALSE}
project_root <- rprojroot::find_root(rprojroot::has_file("README.md"))
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

## Introduction

This report explores the NOAA **HURDAT2** Atlantic-basin hurricane dataset.
HURDAT2 is the official best-track archive for tropical cyclones in the
North Atlantic; the version used here covers 1851–2025 and contains roughly
2,000 storms and 56,000 track points. Each row is a single six-hour
observation of a storm: position, intensity, and (for landfalling points) a
marker. We use it to answer four broad questions:

1. What does the dataset look like in aggregate?
2. What does the spatial distribution of storm tracks look like?
3. Which storms have passed close to a chosen reference city?
4. How have storm frequency, intensity, and season length changed over time?

The analysis is an updated, cleaned-up version of a 2023 coursework project.
A few bugs in the original draft (a latitude-bounds check that was silently
dead code, an unvectorised distance function being called on a vector, a
malformed loop accumulator) have been fixed; the methodology is otherwise
unchanged.

## Data

Daily observations are loaded from a tidy snapshot produced by
`data/fetch_data.R`. The script downloads the current HURDAT2 file from
`nhc.noaa.gov`, parses the fixed-width format, and saves a tidy data frame
called `hurrs` with one row per observation:

```{r load-data}
library(maps); library(tidyverse)
load("data/hurdat2_tidy.RData")

hurrs <- hurrs %>%
  mutate(decade = 10 * (year %/% 10))   # used in Q4
```

```{r summary-table}
summary_stats <- data.frame(
  Statistic = c("Storms", "Track points", "First year", "Last year",
                "Landfall points", "Observations with wind data"),
  Value = c(n_distinct(hurrs$id),
            nrow(hurrs),
            min(hurrs$year),
            max(hurrs$year),
            sum(hurrs$isLandfall),
            sum(!is.na(hurrs$windspeed)))
)
knitr::kable(summary_stats, col.names = c("Statistic", "Value"),
             caption = "HURDAT2 dataset summary")
```

## Part 1: dataset summaries

We can pull out, for example, the full track of Hurricane Katrina (2005):

```{r katrina-rows, echo=TRUE}
katrina <- hurrs %>% filter(id == "AL122005")
cat(sprintf("Katrina (AL122005) track points: %d\n", nrow(katrina)))
```

A per-storm summary function is useful for describing any subset:

```{r summary-fn, echo=TRUE}
Hurdat2_summary <- function(x) {
  list(
    storms       = n_distinct(x$id),
    year_span    = max(x$year) - min(x$year),
    track_points = x %>% group_by(id) %>% summarise(points = n())
  )
}
```

```{r trackpoint-hist, fig.cap="Distribution of track-point counts per storm. The long right tail reflects the longest-lived storms."}
trackpoint <- hurrs %>% group_by(id) %>% summarise(n = n())
ggplot(trackpoint, aes(n)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(x = "Track points per storm", y = "Number of storms") +
  theme_minimal()
```

Storms are named from 1950 onward:

```{r named-start, echo=TRUE}
hurrs %>% filter(name != "UNNAMED") %>% arrange(year) %>% head(1) %>% select(id, name, year)
```

## Part 2: spatial visualisation

A reusable basemap for the Atlantic basin, with overlay tracks coloured by
windspeed:

```{r basemap-fn, echo=TRUE}
hurricaneBasemap <- function() {
  world <- map_data("world")
  ggplot() +
    coord_fixed(xlim = c(-130, 30), ylim = c(0, 90)) +
    geom_polygon(data = world, aes(x = long, y = lat, group = group),
                 color = "gray50", fill = "white")
}
```

```{r all-tracks, fig.cap="All recorded Atlantic storm tracks, coloured by windspeed. Pre-1900 storms lack wind data and appear grey.", fig.height=4.8}
hurricaneBasemap() +
  geom_path(data = hurrs,
            aes(x = longitude, y = latitude, group = id, color = windspeed),
            na.rm = TRUE, alpha = 0.6) +
  scale_color_distiller(type = "seq", direction = 1, palette = "YlOrRd",
                        na.value = "grey80") +
  labs(color = "Wind (kt)") +
  theme(legend.position = "right")
```

A single-year view (Hurricane Andrew and its peers in 1992):

```{r year-1992, fig.cap="Storm tracks in 1992, coloured by windspeed."}
hurricanes_year <- function(x) {
  yearset <- hurrs %>% filter(year == x)
  hurricaneBasemap() +
    geom_path(data = yearset,
              aes(x = longitude, y = latitude, group = id, color = windspeed),
              na.rm = TRUE) +
    scale_color_distiller(type = "seq", direction = 1, palette = "YlOrRd",
                          na.value = "grey80") +
    labs(color = "Wind (kt)")
}
print(hurricanes_year(1992))
```

## Part 3: storms near a reference city

To make distance questions tractable we use an equirectangular approximation
of great-circle distance. Because the function is called inside a `mutate`
on a 56,000-row data frame, it has to be vectorised — the original draft
called a scalar implementation on a vector and silently mis-behaved.

```{r distance-fn, echo=TRUE}
findD <- function(xx, yy, x, y) {
  dy <- yy - y
  dx <- (xx - x) * cos((pi / 180) * ((y + yy) / 2))
  111.325 * sqrt(dx^2 + dy^2)
}

findD_city <- function(x, y, city) {
  ref <- world.cities %>% filter(name == city) %>% slice(1)
  if (nrow(ref) == 0) stop("city not found: ", city)
  findD(ref$long, ref$lat, x, y)
}

# v2 adds a sanity check for swapped lon/lat inputs. The original draft had
# `if (y <-90 & y> 90)` which R parses as `y <- 90; (90 > 90)` — a no-op.
findD_city_v2 <- function(x, y, city) {
  if (any(y < -90 | y > 90, na.rm = TRUE)) {
    warning("Latitude outside [-90, 90]: check for swapped lon/lat.")
  }
  findD_city(x, y, city)
}
```

Which storms have passed within 100 km of Port-au-Prince?

```{r port-au-prince, fig.cap="Storm tracks that have passed within 100 km of Port-au-Prince.", fig.height=4.5}
close <- hurrs %>%
  mutate(distance_to_Prince = findD_city_v2(longitude, latitude, "Port-au-Prince"))
real_close <- close %>% filter(distance_to_Prince <= 100)
hurrs_prince <- hurrs %>% filter(id %in% unique(real_close$id))

port <- world.cities %>% filter(name == "Port-au-Prince")
ggplot() +
  coord_fixed(xlim = c(-100, -25), ylim = c(0, 75)) +
  geom_polygon(data = map_data("world"),
               aes(x = long, y = lat, group = group),
               color = "gray50", fill = "white") +
  geom_path(data = hurrs_prince,
            aes(x = longitude, y = latitude, color = id),
            na.rm = TRUE) +
  geom_point(data = port, aes(x = long, y = lat),
             size = 3, shape = 23, fill = "darkred") +
  geom_text(data = port, aes(x = long, y = lat + 2, label = name), size = 3) +
  theme(legend.position = "none") +
  labs(title = sprintf("Storms within 100 km of Port-au-Prince (%d storms)",
                       n_distinct(hurrs_prince$id)))
```

## Part 4: temporal patterns

### Storms per year

```{r storms-per-year, fig.cap="Distinct Atlantic storms recorded per year. Reporting density is much lower before the satellite era (post-1960s)."}
storms_per_year <- hurrs %>% group_by(year) %>% summarise(count = n_distinct(id))
ggplot(storms_per_year, aes(year, count)) +
  geom_line(color = "steelblue") +
  geom_point(size = 0.7) +
  labs(x = NULL, y = "Distinct storms") +
  theme_minimal()
```

The steep rise in the second half of the 20th century is at least partly an
artefact of better observation — satellites began tracking the open Atlantic
in the 1960s, so storms that would have gone unrecorded earlier now get a
best-track entry.

### Landfall zone over time

```{r landfall-zone, fig.cap="Landfall tracks coloured by year. The set of coastlines affected has expanded poleward and westward over time.", fig.height=4.5}
landed <- hurrs %>% filter(isLandfall)
ggplot() +
  coord_fixed(xlim = c(min(landed$longitude), max(landed$longitude)),
              ylim = c(min(landed$latitude),  max(landed$latitude))) +
  geom_polygon(data = map_data("world"),
               aes(x = long, y = lat, group = group),
               color = "gray50", fill = "white") +
  geom_path(data = landed,
            aes(x = longitude, y = latitude, group = id, color = year)) +
  scale_color_distiller(type = "div", direction = 0, palette = "RdYlBu") +
  labs(color = "Year", title = "Landfall tracks by year") +
  theme_minimal()
```

### Peak windspeed per year

```{r peak-wind, fig.cap="Annual peak windspeed. The trend is upward, but part of the pre-1950 climb reflects the gradual filling-in of missing wind reports rather than a true physical change."}
peak_windspeed <- hurrs %>%
  filter(!is.na(windspeed)) %>%
  group_by(year) %>%
  summarise(peak = max(windspeed))
ggplot(peak_windspeed, aes(year, peak)) +
  geom_line(color = "firebrick") +
  labs(x = NULL, y = "Peak windspeed (kt)") +
  theme_minimal()
```

### Storms per month (the hurricane season)

```{r storms-per-month, fig.cap="Distinct storms per calendar month. September is the peak of the Atlantic hurricane season."}
storms_per_month <- hurrs %>% group_by(month) %>% summarise(count = n_distinct(id))
ggplot(storms_per_month, aes(month, count)) +
  geom_col(fill = "steelblue") +
  scale_x_continuous(breaks = 1:12,
                     labels = month.abb) +
  labs(x = NULL, y = "Distinct storms") +
  theme_minimal()
```

### Season expansion across decades

Stacking each year's storms by month shows the season widening over time:
recent decades have seen off-season storms (January, March, April) that were
essentially absent in the 19th-century record.

```{r season-stack, fig.cap="Storms per year, stacked by month. The coloured bands broaden into May and November-December over time."}
hurrs_year_month <- hurrs %>% group_by(year, month) %>% distinct(id)
ggplot(hurrs_year_month, aes(x = year, fill = factor(month))) +
  geom_histogram(binwidth = 1) +
  scale_fill_brewer(type = "div", direction = 1, palette = "PRGn") +
  labs(x = NULL, y = "Distinct storms", fill = "Month") +
  theme_minimal()
```

## Discussion

A few takeaways:

* HURDAT2 is enormous but inhomogeneous: reporting density changes sharply
  across eras. Any trend in storm counts or peak windspeed has to be read
  against that backdrop.
* The Atlantic hurricane season is clearly concentrated in September, but
  the season appears to have widened over the past several decades — more
  storms recorded in the shoulder months of May and November-December.
* The landfall footprint has expanded poleward and westward compared with
  the early record. Some of that is observation; some is consistent with
  research on poleward migration of tropical-cyclone maximum intensity.

## Reproducibility

Refresh the data with `Rscript data/fetch_data.R`. Re-render the report with
`rmarkdown::render("report/hurricane-analysis.Rmd")`.
